---
title: "HW3 Question 2"
output: html_document
date: '2023-02-12'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load libraries
rm(list=ls())
library(ggplot2)
library(gridExtra)
library(tidyverse)
library(readxl)
library(xts)
library(dlm)
```

```{r}
# Load data and reformat
dlmdata <- read_excel("~/MSDS/predictive models/DLM_Data.xlsx")
dlmdata_missing <- dlmdata[dlmdata$yt == "NA", ] #removing NA values
dlmdata <- dlmdata[dlmdata$yt != "NA", ] #removing NA values
dlmdata$yt <- as.integer(dlmdata$yt) #changing yt from chr to int

# Create plot with given y values
gg_sim <- ggplot(dlmdata, aes(x = time, y = yt)) +
  geom_point(size = 1, color = "seagreen") +
  geom_line(color = "darkseagreen3") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5)) +
  labs(title = 'DLM Data Time Series') 
```

### Part A

The plot is below. The numerical values are as follows: a~40~=3.1155 and R~40~=5.9510.

```{r, fig.align = 'center'}
# Simulate both observed and unobserved from univariate DLM 
# Set true values of unknown parameters
sigma2v_tr <- 9 
sigma2w_tr <- 4
m0_tr <- 0 # Mean of theta
C0_tr <- 25 # Var of theta

# Construct a dlm model object with parameters fixed at their true values
dlm_mod <- dlm(FF = 1.2,
               GG = 0.8,
               V = sigma2v_tr,
               W = sigma2w_tr,
               m0 = m0_tr,
               C0 = C0_tr)   

# Filter the simulated data y using the dlm
dlmdata_filtered <- dlmFilter(y = dlmdata$yt,
                               mod = dlm_mod)

# Store and plot the one-step-ahead predictions of theta and standard errors
dlmdata$pred <- dlmdata_filtered$a
dlmdata$pSE <- sqrt(unlist(
  dlmSvd2var(dlmdata_filtered$U.R, 
             dlmdata_filtered$D.R)))

# Adding one step ahead predictions of theta[t] w/ std errors (in red)
gg_sim +
  geom_line(data = dlmdata, 
            aes(y = pred,
                x = time),
            color = "red",
            size = 1.2) +
  geom_ribbon(data = dlmdata,
              aes(x = time, 
                  ymin = pred - 1.96 * pSE,
                  ymax = pred + 1.96 * pSE),
              fill = "red",
              alpha = 0.2) +
  labs(title = expression(
    paste("One-Step-Ahead Predictions of ",
          theta[t],
          " w/ Standard Errors"))) +
  ylab("theta[t]")

# Predicted mean and variance of theta at time=40
dlmdata$pred[dlmdata$time == 40] # Mean aka at
dlmdata$pSE[dlmdata$time == 40]**2 # Variance aka Rt

```

### Part B

The plot is below. The numerical values are as follows: f~40~=3.7386 and Q~40~=17.5694.

```{r, fig.align = 'center'}
# Create the yt variable column
dlmdata$pred_yt <- rep(0, times=nrow(dlmdata))
for (n in 1:(nrow(dlmdata))){
  dlmdata$pred_yt[n] = dlmdata$pred[n]*1.2
}

# Adding one step ahead predictions of y[t] w/ std errors (in green)
gg_sim +
  geom_line(data = dlmdata, 
            aes(y = pred_yt,
                x = time),
            color = "seagreen",
            size = 1.2) +
  geom_ribbon(data = dlmdata,
              aes(x = time, 
                  ymin = pred_yt - 1.96 * sqrt((pSE**2)*(1.2**2)+(9)),
                  ymax = pred_yt + 1.96 * sqrt((pSE**2)*(1.2**2)+(9))),
              fill = "seagreen",
              alpha = 0.2) +
  labs(title = expression(
    paste("One-Step-Ahead Predictions of ",
          y[t],
          " w/ Standard Errors"))) +
  ylab("y[t]")

# Predicted mean and variance of y at time=40
a = dlmdata$pred[dlmdata$time == 40]
R = dlmdata$pSE[dlmdata$time == 40]**2 
a*1.2 # Mean aka ft
((1.2**2)*R)+9 # Variance aka Qt
```

### Part C

The plot is below. The numerical values are as follows: m~40~=0.3766 and C~40~=3.0484.

```{r, fig.align = 'center'}
# Store and plot the filtering distribution of theta
dlmdata$filtered <- dropFirst(dlmdata_filtered$m)
dlmdata$fSE <- dropFirst(sqrt(unlist(
  dlmSvd2var(dlmdata_filtered$U.C, 
             dlmdata_filtered$D.C))))

gg_sim +
  geom_line(data = dlmdata, 
            aes(y = pred,
                x = time),
            color = "red",
            size = .5) +
  geom_line(data = dlmdata, 
            aes(y = filtered,
                x = time),
            color = "blue",
            size = 1.2) +
  geom_ribbon(data = dlmdata,
              aes(x = time, 
                  ymin = filtered - 1.96 * fSE,
                  ymax = filtered + 1.96 * fSE),
              fill = "blue",
              alpha = 0.2) +
  labs(title = expression(
    paste("Mean of the filtering distribution of ",
          theta[t],
          " w/ Standard Errors"))) +
  ylab("theta[t]")


# Predicted mean and variance of y at time=40
dlmdata$filtered[dlmdata$time == 40] # Mean aka mt
dlmdata$fSE[dlmdata$time == 40]**2 # Variance aka Ct
```

### Part D

The recursive manual Kalman Filtering process is shown below.

```{r}
# Starting Tt|yt at time=22 and constants
m=3.539
C=3.048
G=0.8
F=1.2
w=4
v=9
e=0 # Setting the forecast error to 0 since we don't have yts

# Looping through Kalman Filter
for (n in 1:6) {
# Step 1 Kalman Filter (Tt|yt-1)
a=G*m
R=((G**2)*C)+w

# Step 2 Kalman Filter (yt|yt-1)
f=F*a
Q=((F**2)*R)+v

# Step 3 Kalman Filter (Tt|yt)
m=a+(F*R*(1/Q)*e)
C=R-((R**2)*(F**2)*(1/Q))
}

a
R
```

### Part E

The plot and values for $\theta$~t~ such that y~t~ is missing are show below. The missing values are repeated here for ease:

* Time=11, $\theta$=1.950

* Time=23, $\theta$=0.492

* Time=24, $\theta$=-0.342

* Time=25, $\theta$=-0.800

* Time=26, $\theta$=-1.940

* Time=27, $\theta$=-0.724

* Time=64, $\theta$=1.960

* Time=80, $\theta$=-2.24

```{r, fig.align = 'center'}
# Store and plot the smoothed predictions of theta and standard errors
dlmdata_smoothed <- dlmSmooth(dlmdata_filtered)
dlmdata$smoothed <- dropFirst(dlmdata_smoothed$s)
dlmdata$sSE <- dropFirst(sqrt(unlist(
  dlmSvd2var(dlmdata_smoothed$U.S, 
             dlmdata_smoothed$D.S))))

gg_sim +
  geom_line(data = dlmdata, 
            aes(y = pred,
                x = time),
            color = "red",
            size = .5) +
  geom_line(data = dlmdata, 
            aes(y = filtered,
                x = time),
            color = "blue",
            size = .5) +
  geom_line(data = dlmdata, 
            aes(y = smoothed,
                x = time),
            color = "darkgreen",
            size = 1.2) +
  geom_ribbon(data = dlmdata,
              aes(x = time, 
                  ymin = smoothed - 2 * sSE,
                  ymax = smoothed + 2 * sSE),
              fill = "green",
              alpha = 0.2) +
  labs(title = expression(
    paste("Smoothed Values of ",
          theta[t],
          " w/ Standard Errors"))) +
  ylab("theta[t]")

# Creating a  vector to store the NA times 
missing_times <- c()

# Append t+1 polio counts to NA times
for (n in dlmdata_missing$time){
  missing_times <- append(missing_times, dlmdata_smoothed$s[n])}

# Separating only missing NA times w/ smothed estimates
dlmdata_missing$new_thetat <- missing_times
dlmdata_missing
```

### Part F

The plot is shown below. 

Earlier in forecasting for future values of y, there is more certainty about what the value will be. As you go further out, i.e., at time 110 vs time 101, it is less certainty that the data will continue to follow the same trend, meaning that there is a lower predictive variance for y~101~ (17.5694) than y~110~ (24.8661).

```{r, fig.align = 'center'}
# DLM forecast function
forecast_mod <- dlmForecast(dlmdata_filtered, nAhead = 10)

# Exctracting standard deviation and appending mean/var to dlmdata
dlmdata$ytstdev <- rep(0, times=nrow(dlmdata))

for (n in 1:(nrow(forecast_mod$f))){
  new_row <- c(forecast_mod$f[n], 100+n, forecast_mod$f[n], "","","","","","")
  dlmdata <- rbind(dlmdata, new_row) 
  dlmdata$ytstdev[dlmdata$time == 100+n] = as.numeric(forecast_mod$Q[n])
}


# Changing yt/ytstdev from chr to int
dlmdata$yt <- as.integer(dlmdata$yt) 
dlmdata$ytstdev <- as.integer(dlmdata$ytstdev)
dlmdata$time <- as.integer(dlmdata$time)

gg_sim <- ggplot(dlmdata, aes(x = time, y = yt)) +
  geom_point(size = 1, color = "seagreen") +
  geom_line(color = "darkseagreen3") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5)) +
  labs(title = 'DLM Data Time Series') 

gg_sim +
  geom_ribbon(data = dlmdata,
              aes(x = time, 
                  ymin = yt - 1.96 * sqrt(ytstdev),
                  ymax = yt + 1.96 * sqrt(ytstdev)),
              fill = "seagreen",
              alpha = 0.2) +
  labs(title = expression(
    paste("Original ", y[1:100]," Values with 10 Future Values"))) +
  ylab("y")

forecast_mod$Q[1] #Q101
forecast_mod$Q[10] #Q110
```