---
title: "HW3 Question 1"
output: html_document
date: '2023-02-10'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Setup
rm(list=ls())
library(ggplot2)
library(gridExtra)
library(tidyverse)
library(xts)
library(depmixS4)
library(dlm)
library(gamlss.data)

# DF setup with polio counts in one column
df <- data.frame(polio, date=paste(month.abb[cycle(polio)], sep=" ", trunc(time(polio))))
```

### Part A

The model is below. The AIC is 530.0655. The $\lambda$~1~=0.7902 and the $\lambda$~2~=4.1779.The probabilities of the state transitional matrix are as follows:

* State 1 to State 1: 93.2%

* State 1 to State 2:  6.8%

* State 2 to State 1:  3.3%

* State 2 to State 2: 67.0%


```{r, fig.align = 'center'}
# Create the non-stationary HMM 
eq_mod <- depmix(polio ~ 1, # this is the intercept for poisson paramater
                 data = df,
                 family = poisson(),
                 nstates = 2, # 2 state HMM
                 ntimes = nrow(df))


# Fit the model using an EM algorithm
set.seed(900)
eq_fit <- fit(eq_mod)
summary(eq_fit) #lambda for each hidden state are the st1 and st2 response params
eq_fit

# Get estimate of lambda
getpars(eq_fit)

# Exponentiate to get back on the poisson scale
exp(getpars(eq_fit)[7:8])

# Get the posterior distribution of the states from best guess of latent states at each time points
df$states <- as.factor(posterior(eq_fit, type = "viterbi")$state)

# Adding a time column to df so it plots in order vs alphabetical
time = 1:nrow(df)
df$time = time
# Adding more readable date labels for x axis
jan_months <- c()
jan_labels <- c()
for (n in 0:13){
  jan_months <- append(jan_months, 12*n+1)
  jan_labels <- append(jan_labels, paste("Jan", 1970+n))
}

# Plotting the time series
ggplot(df, aes(x = time, y = polio)) +
  geom_point(size = 1, color = "seagreen") +
  geom_text(aes(label = states, color = states), nudge_y = 0.7) +
  geom_line(color = "darkseagreen3") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5), axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  scale_x_continuous(breaks = jan_months, labels = jan_labels) +
  labs(title = '2 State Model:\n Polio Counts from 1970 to 1983') + 
  scale_colour_brewer(palette = "Set1")
```

### Part B

For data points most likely under State 1, specifically when $\lambda$~1~=0.7902, the number of polio cases is lower (closer to 0.7902) with less variability from this point (variance=0.7902). For data points most likely under State 2, specifically when $\lambda$~2~=4.1779, the number of polio cases is higher (closer to 4.1779) with more variability from this point (variance=4.1779).

If we look at points with 4+ polio counts, we are more likely to observe exactly 0 cases (38.46% chance) following this as opposed to 2+ (23.08% chance). This is not consistent with basic knowledge about infectious disease, given that following high polio counts, it is most likely that polio counts would remain high or slowly drop off as opposed to immediately going to 0. 

```{r}
# Subsetting the polio data frame for at least 4 cases, store the time t polio counts
highpolio <- df[df$polio > '3', ]
# Creating an empty vector to store the time t+1 polio counts
after_highpolio <- c()
# Append t+1 polio counts
for (n in highpolio$time){
  after_highpolio <- append(after_highpolio, df$polio[df$time==(n+1)])
}
# Percentage of more than 2 polio counts after at least 4 polio
length(after_highpolio[after_highpolio>2])/length(after_highpolio)
# Percentage of 0 polio counts after at least 4 polio
length(after_highpolio[after_highpolio==0])/length(after_highpolio)
```

### Part C

The model is below. The AIC is 529.9555. The $\lambda$~1~=2.2980, the $\lambda$~2~=0.6487, and $\lambda$~3~=8.2571.The probabilities of the state transitional matrix are as follows:

* State 1 to State 1: 69.6%

* State 1 to State 2: 21.3%

* State 1 to State 3:  9.1%

* State 2 to State 1:  9.3%

* State 2 to State 2: 90.7%

* State 2 to State 3:  0.0%

* State 3 to State 1: 61.4%

* State 3 to State 2: 10.6%

* State 3 to State 3: 28.0%

```{r, fig.align = 'center'}
# New dataframe without previous states
df <- data.frame(polio, date=paste(month.abb[cycle(polio)], sep=" ", trunc(time(polio))))

# Create the non-stationary HMM 
eq_mod <- depmix(polio ~ 1, # this is the intercept for poisson paramater
                 data = df,
                 family = poisson(),
                 nstates = 3, # 3 state HMM
                 ntimes = nrow(df))


# Fit the model using an EM algorithm
set.seed(900)
eq_fit <- fit(eq_mod)
summary(eq_fit) #lambda for each hidden state are the st1 and st2 response params
eq_fit

# Get estimate of lambda
getpars(eq_fit)

# Exponentiate to get back on the poisson scale
exp(getpars(eq_fit)[13:15])

# Get the posterior distribution of the states from best guess of latent states at each time points
df$states <- as.factor(posterior(eq_fit, type = "viterbi")$state)

# Adding a time column to df so it plots in order vs alphabetical
df$time = time

# Plotting the time series
ggplot(df, aes(x = time, y = polio)) +
  geom_point(size = 1, color = "seagreen") +
  geom_text(aes(label = states, color = states), nudge_y = 0.7) +
  geom_line(color = "darkseagreen3") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5), axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  scale_x_continuous(breaks = jan_months, labels = jan_labels) +
  labs(title = '3 State Model:\n Polio Counts from 1970 to 1983') + 
  scale_colour_brewer(palette = "Set1")
```
